On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-04-13 manuscripts/PNAS/chazthm_pnas.tex
sml edits removed all comments

\significancestatement{Many naturally occurring signals (visual scenery, speech, EEG, etc.) lacking domain-specific formal models can nonetheless be usefully characterized as linear combinations of few elementary waveforms drawn from a large `dictionary'. We give general conditions guaranteeing when such dictionaries are uniquely and stably determined by data. The result justifies the use of this model as a constraint for blind source separation, and may help explain the observed universality of emergent representations in sparse coding models of some natural phenomena.
%These results provide theoretical grounding for  the application of dictionary learning to recovering sparse signals from unknown noisy compressive measurements.
}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{B}lind source separation is a classical problem in signal processing \cite{sato1975method}.
A common modern assumption is that each of $N$ observed $n$-dimensional signals is a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary'' of size $m$, typically with $k < m \ll N$ (see \cite{Zhang15} for a comprehensive review of this and related models).
%A popular underlying assumption is that there exists a `dictionary' of $m$ elementary waveforms, at most $k$ of which need be linearly combined to represent each signal in a given dataset of size $N$, typically where $k < m \ll N$.  
Approximating solutions to this sparsity-constrained inverse problem have provided insight into the structure of many signal classes lacking domain-specific formal models (e.g., in vision \cite{wang2015sparse}).  In particular, it has been shown that response properties of simple-cell neurons in mammalian visual cortex emerge from optimizing a dictionary to represent small patches of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent}, a major advance in computational neuroscience. A curious aspect of this finding is that the latent waveforms (e.g. `Gabor' wavelets) estimated from data appear to be canonical \cite{donoho2001can};
i.e., they are found in learned dictionaries independent of algorithmic implementation or natural image training set.

% (e.g., Fourier bases, wavelets, etc.)

Motivated by these discoveries and earlier work in the theory of neural communication \cite{Coulter10, Isely10}, we address when dictionaries and the sparse representations they induce are uniquely determined by data.  Answers to this question also have other real-world implications.  For example, a sparse coding analysis of local painting style can be used for forgery detection \cite{hughes2010, Olshausen10}, but only if all dictionaries consistent with training data do not differ appreciably in their ability to sparsely encode new samples. %the identifying sparse components of the artwork can be consistently estimated. % are independent of implementation idiosyncrasies. 
Fortunately, algorithms with proven recovery of generating dictionaries under certain 
conditions have recently been proposed (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art). Few theorems, however, can be cited to explain this uniqueness phenomenon more generally.

Here, we prove very generally that uniqueness and stability in sparse linear coding is an expected property of the model. %; that is, it is a well-posed problem (in the sense of Hadamard \cite{Hadamard1902}). 
%In other words, whenever enough sparsely represented data generated from a dictionary (satisfying a spark condition) are observed, the original codes and dictionary are uniquely determined up to an error that is commensurate with the measurement noise. 
More specifically, dictionaries that preserve sparse codes (i.e., satisfy a `spark condition') are identifiable from as few as \mbox{$N = m(k-1){m \choose k} + m$} noisy sparse linear combinations of their columns up to an error linear in the noise (Thm.~\ref{DeterministicUniquenessTheorem}). In fact, provided $n \geq \min(2k,m)$, in almost all cases the dictionary learning problem is well-posed (as per Hadamard \cite{Hadamard1902}) given enough data (Cor.~\ref{ProbabilisticCor}). Moreover, these guarantees hold without assuming the recovered matrix satisfies a spark condition, even when the number $m$ of dictionary elements is unknown. The explicit, algorithm-independent criteria we provide should be a useful tool in the theory of sparse representations.

%Other applications include: 1) a sufficient diversity of sparse codes sent through a random linear compressive channel are identifiable and 2) an explanation for universal representation of natural signals in sparse coding models from neuroscience.

@@

Note that any particular solution $(\mathbf{B}, \mathbf{\bar x}_1 \ldots, \mathbf{\bar x}_N)$ to this problem gives rise to an orbit of equivalent solutions $(\mathbf{BPD}, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_N)$, where $\mathbf{P}$ is any permutation matrix and $\mathbf{D}$ any invertible diagonal. Previous theoretical work addressing the noiseless case $\eta =0$ (e.g., \cite{li2004analysis, Georgiev05, Aharon06, Hillar15}) has shown that a solution to Prob.~\ref{InverseProblem} (when it exists) is indeed unique up to this inherent ambiguity provided the $\mathbf{x}_i$ are sufficiently diverse and the generating matrix $\mathbf{A}$ satisfies the \textit{spark condition} from compressive sensing (CS):
\begin{align}\label{SparkCondition}
\mathbf{A}\mathbf{x}_1 = \mathbf{A}\mathbf{x}_2 \implies \mathbf{x}_1 = \mathbf{x}_2, \ \ \ \text{for all $k$-sparse } \mathbf{x}_1, \mathbf{x}_2,
\end{align}
%
which would be, in any case, necessary for uniqueness. Our concern here is the stability of these solutions with respect to noise in the data.

\begin{definition}\label{maindef}
Fix $Y = \{ \mathbf{y}_1, \ldots, \mathbf{y}_N\} \subset \mathbb{R}^n$. We say $Y$ has a \textbf{$k$-sparse representation in $\mathbb{R}^m$} if there exists a matrix $\mathbf{A}$ and $k$-sparse $\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$ such that $\mathbf{y}_i = \mathbf{A}\mathbf{x}_i$ for all $i$. %Equivalently, $\mathbf{Y} = \mathbf{A}\mathbf{X}$ for $\mathbf{Y}$ and $\mathbf{X}$ having the $\mathbf{y}_i$ and $\mathbf{x}_i$ as their columns, respectively. 
This representation is \textbf{stable} if for every $\delta_1, \delta_2 \geq 0$, there exists $\varepsilon(\delta_1, \delta_2) \geq 0$ (with $\varepsilon > 0$ when  $\delta_1, \delta_2 > 0$) such that if $\mathbf{B}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^m$ satisfy:
\begin{align*}
\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon,\ \ \   \text{for all $i$},
\end{align*}
%
then there is some permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ such that for all $i, j$:
\begin{align}\label{def1}
\|\mathbf{A}_j - (\mathbf{BPD})_j\|_2 \leq \delta_1 \ \ \text{and} \ \ \|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_i\|_1 \leq \delta_2.
\end{align}
\end{definition}

To see how Def. \ref{maindef} directly relates to Prob. \ref{InverseProblem}, suppose that $Y$
% \mbox{$Y = \{\mathbf{A} \mathbf{x}_1, \ldots, \mathbf{A}\mathbf{x}_N\}$} 
has a stable $k$-sparse representation in $\mathbb{R}^m$ and fix $\delta_1, \delta_2$ to be the desired recovery accuracy in \eqref{def1}. Consider now any dataset $Z$ generated as in \eqref{LinearModel} with $\eta \leq \frac{1}{2} \varepsilon(\delta_1, \delta_2)$. Then from the triangle inequality, it follows that any matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^m$ solving Prob.~\ref{InverseProblem} are necessarily close to the original dictionary $\mathbf{A}$ and sparse codes $\mathbf{x}_i$.

%Our main technical finding is that a sufficient quantity of measurements generated from any dictionary satisfying the spark condition has a stable sparse representation.  More specifically, such dictionaries are identifiable from as few as \mbox{$N = m(k-1){m \choose k} + m$} sparse linear combinations of their columns up to an error that is linear in the measurement noise (Thm.~\ref{DeterministicUniquenessTheorem}). In fact, provided $n \geq k$, in almost all cases the dictionary learning problem is well-posed (in the sense of Hadamard \cite{Hadamard1902}) given enough data (Cor.~\ref{ProbabilisticCor}). Importantly, these guarantees hold without assuming the recovered matrix satisfies a spark condition. Our results  also apply when only an upper bound on the number of dictionary elements is known. The explicit, algorithm-independent criteria we provide should be a useful tool in the theory of dictionary learning.

@@

Before stating our main results, we first identify combinatorial criteria on the support sets of generating codes that imply stable sparse representations.  Letting $\{1, \ldots, m\}$ be denoted $[m]$, its power set $2^{[m]}$, and ${[m] \choose k}$ the set of subsets of $[m]$ of size $k$,
% Let $\text{\rm Span}\{\mathbf{v}_1, \ldots, \mathbf{v}_\ell\}$ be the $\mathbb{R}$-linear span of vectors $\mathbf{v}_1, \ldots, \mathbf{v}_\ell$. 
we say a hypergraph $E \subseteq 2^{[m]}$ on vertices $[m]$ is \textit{$k$-uniform} when in fact $E \subseteq {[m] \choose k}$. We also say $E$ is \emph{regular} when every element of $[m]$ is contained in exactly $\ell$ elements of $E$ for some $\ell > 0$ (for given $\ell$, we say $E$ is \textit{$\ell$-regular}).

@@

It is easy to verify that for every $k < m$, there is a regular $k$-uniform hypergraph that satisfies the SIP; for instance, consecutive intervals of length $k$ in some cyclic order on $[m]$.
%\begin{proposition}
%For every $k < m$, there is a regular $k$-uniform hypergraph that satisfies the singleton intersection property.
%\end{proposition}

%The second ingredient necessary for the statement of Thm.~\ref{DeterministicUniquenessTheorem} 
Next, we describe a quantitative version of the spark condition.  % We first explain how the spark condition \eqref{SparkCondition} relates to t
The \emph{lower bound} \cite{Grcar10} of a matrix $\mathbf{M} \in \mathbb R^{n \times m}$ is the largest number $\alpha$ such that \mbox{$\|\mathbf{M}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2$} for all $\mathbf{x} \in \mathbb{R}^m$. By compactness of the unit sphere, every injective linear map has a nonzero lower bound; hence, if $\mathbf{M}$ satisfies \eqref{SparkCondition}, then each submatrix formed from $2k$ of its columns or less has a nonzero lower bound. This motivates the following definition: % We therefore define the following domain-restricted lower bound:
\begin{align*}
L_k(\mathbf{M}) := \frac{1}{\sqrt{k}}\max \{ \alpha : \|\mathbf{M}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2 \text{ for all $k$-sparse } \mathbf{x}\}.
\end{align*}

%Clearly, $\sqrt{k} L_k(\mathbf{A}) \geq \sqrt{k'}L_{k'}(\mathbf{A})$ whenever $k < k'$, and 
Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$. We note that the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the CS literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.%, Foucart2009}.

@@

\begin{theorem}\label{DeterministicUniquenessTheorem}
%Fix integers $n, \ell$, and $k < m$. 
Let $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfy \eqref{SparkCondition} for $k < m$, and suppose for some $\bar m$ that more than \mbox{$(k-1){\bar m \choose k}$} $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} are supported in general linear position on each set in an $\ell$-regular $E \subseteq {[m] \choose k}$ with the SIP.  There is a constant\footnote{We delay defining the constant $C_1$ until Section \ref{DUT} (\eqref{Cdef1}).} $C_1 > 0$ for which the following holds for $\varepsilon < L_2(\mathbf{A}) / C_1$:

@@

%\begin{theorem}\label{DeterministicUniquenessTheorem}
%Fix integers $n, \ell$, and $k < m$. Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition} and that $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} contain at least \mbox{$(k-1){m \choose k}+1$} vectors supported in general linear position on each set in some $\ell$-regular $E \subseteq {[m] \choose k}$ with the SIP.  There exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$.

%Every matrix $\mathbf{B} \in \mathbb{R}^{n \times m}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^{m}$ with  \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon$} for all $i$ necessarily satisfies:
%\begin{align}\label{Cstable}
%\|\mathbf{A}_j - (\mathbf{B}\mathbf{PD})_j\|_2 \leq C_1 \varepsilon,\ \ \   \text{for all $j \in [m]$},
%\end{align}
%for some permutation matrix $\mathbf{P}$ and invertible diagonal $\mathbf{D}$.

%Furthermore, if $\varepsilon < L_{2k}(\mathbf{A}) / C_1$, then $\mathbf{B}$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$, and:
%\begin{align}\label{b-PDa}
%\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,\ \ \   \text{for all $i \in [N]$}.
%\end{align}
%\end{theorem}

%\begin{theorem}\label{DeterministicUniquenessTheorem}
%Fix integers $n, \ell$, and $k < m \leq \bar m$. Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition} and that $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} contain at least \mbox{$(k-1){\bar m \choose k}+1$} vectors supported in general linear position on each set in some $\ell$-regular $E \subseteq {[m] \choose k}$ satisfying the SIP.  There exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$.
%
%Every matrix $\mathbf{B} \in \mathbb{R}^{n \times \bar m}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^{\bar m}$ with  \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon$} for all $i$ necessarily satisfies:
%\begin{align}\label{Cstable}
%\|\mathbf{A}_{\phi(j)} - (\mathbf{B}_{J}\mathbf{PD})_j\|_2 \leq C_1 \varepsilon,\ \ \   \text{for $j = 1, \ldots, |J|$},
%\end{align}
%%
%for some $J \subseteq [\bar m]$ of size \mbox{$\lfloor \bar m - \ell(\bar m - m) \rfloor$}, injective map $\phi$, permutation matrix $\mathbf{P}$, and invertible diagonal matrix $\mathbf{D}$.
%%\footnote{We note that the condition $\varepsilon < L_2(\mathbf{A})$ above is necessary. When \mbox{$\mathbf{A}$ = $I$} and $\mathbf{x}_i = \mathbf{e}_i$, it turns out that $C_1 = 1$ and there is a $\mathbf{B}$ and $1$-sparse $\mathbf{\bar x}_i$ with $\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon$ violating \eqref{Cstable}.}
%
%Furthermore, if $\varepsilon < L_{2k}(\mathbf{A}) / C_1$, then $\mathbf{B}_J$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}_J\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$, and:
%\begin{align}\label{b-PDa}
%\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,\ \ \   \text{for all $i \in [N]$}.
%\end{align}
%\end{theorem}

To be clear, Thm.~\ref{DeterministicUniquenessTheorem} says that the smaller the difference $\bar m - m$, the more columns and coefficients of the original dictionary $\mathbf{A}$ and codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled dictionary $\mathbf{B}$ and codes $\mathbf{\bar x}_i$. 
%In particular, when $\bar m = m$ all columns of $\mathbf{A}$ and coefficients of each $\mathbf{x}_i$ recoverable, as per Def.~\ref{maindef}. 
The implication in the case $\bar{m} = m$ is precisely that $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with \eqref{def1} guaranteed provided $\varepsilon$ in Def.~\ref{maindef} does not exceed: 
\begin{align}\label{epsdel}
\varepsilon(\delta_1, \delta_2) := \min \left\{ \frac{\delta_1}{ C_1 }, \frac{ \delta_2 L_{2k}(\mathbf{A})}{ 1 + C_1 \left( \max_{i \in [N]} \|\mathbf{x}_i\|_1  + \delta_2 \right) } \right\}.
\end{align}
  % , since it requires additional definitions not critical to the statement of our results.

% Note also that Thm~\ref{DeterminsiticUniquenessTheorem} gives conditions for when a single solution to Problem~\ref{InverseProblem} in fact represents \emph{all} possible solutions, related by transformations of the form $PD$.

@@

%\begin{proposition}
%If $E$ is an $\ell$-regular hypergraph of rank $k$ satisfying the singleton intersection property then $|E| \geq \ell m/k$. 
%\end{proposition}
%Note that every $\ell$-regular hypergraph $E \subseteq {[m] \choose k}$ satisfies: \[|E|k = \sum_{S \in E}|S| = \sum_{i \in [m]} \deg(i) = \ell m.\] 
As already mentioned, it is easy to construct a regular hypergraph $E \subseteq {[m] \choose k}$ with the SIP having cardinality $|E| = m$, implying the lower bound for sample size $N$ from the introduction. In many cases, however, the SIP can be achieved with far fewer supports; for example, when $k = \sqrt{m}$ one can take $E$ to be the $2k$ rows and columns formed by arranging $[m]$ in a square grid.  
%In light of Cor.~\ref{DeterministicUniquenessCorollary}, it would be interesting to determine other such $E$. %[ ** TODO ** can we show for every $k, m$ that there is an $E$ satisfying the SIP with $|E| = \ell m/k$ for every $\ell$ such that $\ell m/k$ is an integer? Then lower bound on $N$ comes from the smallest such $\ell$.]

@@

\begin{corollary}\label{SLCopt}
If all of the assumptions of Thm.\ref{DeterministicUniquenessTheorem} hold with more than $(k-1)\left[ {\bar m \choose k} + |E|k{\bar m \choose k-1}\right]$ vectors $\mathbf{x}_i$ supported on each $S \in E$, then all solutions to Prob.~\ref{OptimizationProblem} necessarily satisfy the implications \eqref{Cstable} and \eqref{b-PDa} of Thm.~\ref{DeterministicUniquenessTheorem}.
\end{corollary}
%\begin{corollary}\label{SLCopt}
%If the assumptions of Thm.\ref{DeterministicUniquenessTheorem} hold with more than $(k-1)\left[ {\bar m \choose k} + |E|k{\bar m \choose k-1}\right]$ vectors $\mathbf{x}_i$ supported on each $S \in E$, then all matrices $\mathbf{B} \in \mathbb{R}^{n \times \bar m}$ and $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^{\bar m}$ solving:
%\begin{align}\label{minsum}
%\min \sum_{i = 1}^N \|\mathbf{\bar x}_i\|_0 \ \ 
%\text{subject to} \ \ \|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon, \ \text{for all $i$},
%\end{align}
%
%necessarily satisfy implications 1) and 2) from Thm.~\ref{DeterministicUniquenessTheorem}.

%must have $\bar m \geq m$ and, provided $p = \bar m - \ell(\bar m - m) > 0$, necessarily satisfy \eqref{Cstable}
%%\begin{align}\label{Cstable}
%%\|(\mathbf{A}_J- \mathbf{B}_{\bar J} \mathbf{PD} )_j\|_2 \leq C_1 \varepsilon, \ \  \text{for  $j \in J$},
%%\end{align}
%for some $J \in {[m] \choose p}$, $\bar J \in {[\bar m] \choose p}$, permutation matrix $\mathbf{P}$, and invertible diagonal matrix $\mathbf{D}$.
%\end{corollary}

>>>>

\significancestatement{Many naturally occurring signals (visual scenery, speech, EEG, etc.) lacking domain-specific formal models can nonetheless be usefully characterized as linear combinations of few elementary waveforms drawn from a large `dictionary'. We give general conditions guaranteeing when such dictionaries are uniquely and stably determined by data. The result justifies the use of this model as a constraint for blind source separation, and may help explain the observed universality of emergent representations in sparse models of some natural phenomena.
}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{B}lind source separation is a classical problem in signal processing \cite{sato1975method}.
A common modern assumption is that each of $N$ observed $n$-dimensional signals is a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary'' of size $m$, typically with $k < m \ll N$ (see \cite{Zhang15} for a comprehensive review of this and related models).
Approximating solutions to this sparsity-constrained inverse problem have provided insight into the structure of many signal classes lacking domain-specific formal models (e.g., in vision \cite{wang2015sparse}).  In particular, it has been shown that response properties of simple-cell neurons in mammalian visual cortex emerge from optimizing a dictionary to represent small patches of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent}, a major advance in computational neuroscience. A curious aspect of this finding is that the latent waveforms (e.g. `Gabor' wavelets) estimated from data appear to be canonical \cite{donoho2001can};
i.e., they are found in learned dictionaries independent of algorithm or natural image training set.

Motivated by these discoveries and earlier work in the theory of neural communication \cite{Coulter10, Isely10}, we address when dictionaries and the sparse representations they induce are uniquely determined by data.  Answers to this question also have other real-world implications.  For example, a sparse coding analysis of local painting style can be used for forgery detection \cite{hughes2010, Olshausen10}, but only if all dictionaries consistent with training data do not differ appreciably in their ability to sparsely encode new samples. 
Fortunately, algorithms with proven recovery of generating dictionaries under certain 
conditions have recently been proposed (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art). Few theorems, however, can be cited to explain this uniqueness phenomenon more generally.

Here, we prove very generally that uniqueness and stability in sparse linear coding is an expected property of the model. 
More specifically, dictionaries that preserve sparse codes (i.e., satisfy a `spark condition') are identifiable from as few as \mbox{$N = m(k-1){m \choose k} + m$} noisy sparse linear combinations of their columns up to an error linear in the noise (Thm.~\ref{DeterministicUniquenessTheorem}). In fact, provided $n \geq \min(2k,m)$, in almost all cases the dictionary learning problem is well-posed (as per Hadamard \cite{Hadamard1902}) given enough data (Cor.~\ref{ProbabilisticCor}). Moreover, these guarantees hold without assuming the recovered matrix satisfies a spark condition, even when the number $m$ of dictionary elements is unknown. The explicit, algorithm-independent criteria we provide should be a useful theoretical tool.  %  in the theory of sparse representations.

@@

Note that any particular solution $(\mathbf{B}, \mathbf{\bar x}_1 \ldots, \mathbf{\bar x}_N)$ to this problem gives rise to an orbit of equivalent solutions $(\mathbf{BPD}, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_N)$, where $\mathbf{P}$ is any permutation matrix and $\mathbf{D}$ any invertible diagonal. Previous theoretical work addressing the noiseless case $\eta =0$ (e.g., \cite{li2004analysis, Georgiev05, Aharon06, Hillar15}) has shown that a solution to Prob.~\ref{InverseProblem} (when it exists) is indeed unique up to this inherent ambiguity provided the $\mathbf{x}_i$ are sufficiently diverse and the generating matrix $\mathbf{A}$ satisfies the \textit{spark condition} from compressive sensing (CS):
\begin{align}\label{SparkCondition}
\mathbf{A}\mathbf{x}_1 = \mathbf{A}\mathbf{x}_2 \implies \mathbf{x}_1 = \mathbf{x}_2, \ \ \ \text{for all $k$-sparse } \mathbf{x}_1, \mathbf{x}_2,
\end{align}
%
which would be, in any case, necessary for uniqueness. Our concern here is solution stability with respect to noise.

\begin{definition}\label{maindef}
Fix $Y = \{ \mathbf{y}_1, \ldots, \mathbf{y}_N\} \subset \mathbb{R}^n$. We say $Y$ has a \textbf{$k$-sparse representation in $\mathbb{R}^m$} if there exists a matrix $\mathbf{A}$ and $k$-sparse $\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$ such that $\mathbf{y}_i = \mathbf{A}\mathbf{x}_i$ for all $i$. 
This representation is \textbf{stable} if for every $\delta_1, \delta_2 \geq 0$, there exists $\varepsilon(\delta_1, \delta_2) \geq 0$ (with $\varepsilon > 0$ when  $\delta_1, \delta_2 > 0$) such that if $\mathbf{B}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^m$ satisfy:
\begin{align*}
\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon,\ \ \   \text{for all $i$},
\end{align*}
%
then there is some permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ such that for all $i, j$:
\begin{align}\label{def1}
\|\mathbf{A}_j - (\mathbf{BPD})_j\|_2 \leq \delta_1 \ \ \text{and} \ \ \|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_i\|_1 \leq \delta_2.
\end{align}
\end{definition}

To see how Def. \ref{maindef} directly relates to Prob. \ref{InverseProblem}, suppose that $Y$

has a stable $k$-sparse representation in $\mathbb{R}^m$ and fix $\delta_1, \delta_2$ to be the desired recovery accuracy in \eqref{def1}. Consider now any dataset $Z$ generated as in \eqref{LinearModel} with $\eta \leq \frac{1}{2} \varepsilon(\delta_1, \delta_2)$. Then from the triangle inequality, it follows that any matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^m$ solving Prob.~\ref{InverseProblem} are necessarily close to the original dictionary $\mathbf{A}$ and sparse codes $\mathbf{x}_i$.

@@

Before stating our main results, we first identify combinatorial criteria on the support sets of generating codes that imply stable sparse representations.  Letting $\{1, \ldots, m\}$ be denoted $[m]$, its power set $2^{[m]}$, and ${[m] \choose k}$ the set of subsets of $[m]$ of size $k$,
we say a hypergraph $E \subseteq 2^{[m]}$ on vertices $[m]$ is \textit{$k$-uniform} when in fact $E \subseteq {[m] \choose k}$. We also say $E$ is \emph{regular} when every element of $[m]$ is contained in exactly $\ell$ elements of $E$ for some $\ell > 0$ (for given $\ell$, we say $E$ is \textit{$\ell$-regular}).

@@

It is easy to verify that for every $k < m$, there is a regular $k$-uniform hypergraph that satisfies the SIP; for instance, consecutive intervals of length $k$ in some cyclic order on $[m]$.

Next, we describe a quantitative version of the spark condition.  
The \emph{lower bound} \cite{Grcar10} of a matrix $\mathbf{M} \in \mathbb R^{n \times m}$ is the largest number $\alpha$ such that \mbox{$\|\mathbf{M}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2$} for all $\mathbf{x} \in \mathbb{R}^m$. By compactness of the unit sphere, every injective linear map has a nonzero lower bound; hence, if $\mathbf{M}$ satisfies \eqref{SparkCondition}, then each submatrix formed from $2k$ of its columns or less has a nonzero lower bound. This motivates the following definition: 
\begin{align*}
L_k(\mathbf{M}) := \frac{1}{\sqrt{k}}\max \{ \alpha : \|\mathbf{M}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2 \text{ for all $k$-sparse } \mathbf{x}\}.
\end{align*}

Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$. We note that the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the CS literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.

@@

\begin{theorem}\label{DeterministicUniquenessTheorem}
Let $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfy \eqref{SparkCondition} for $k < m$, and suppose for some $\bar m$ that more than \mbox{$(k-1){\bar m \choose k}$} $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} are supported in general linear position on each set in an $\ell$-regular $E \subseteq {[m] \choose k}$ with the SIP.  There is a constant\footnote{We delay defining the constant $C_1$ until Section \ref{DUT} (\eqref{Cdef1}).} $C_1 > 0$ for which the following holds for $\varepsilon < L_2(\mathbf{A}) / C_1$:

@@

To be clear, Thm.~\ref{DeterministicUniquenessTheorem} says that the smaller the difference $\bar m - m$, the more columns and coefficients of the original dictionary $\mathbf{A}$ and codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled dictionary $\mathbf{B}$ and codes $\mathbf{\bar x}_i$. 
The implication in the case $\bar{m} = m$ is precisely that $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with \eqref{def1} guaranteed provided $\varepsilon$ in Def.~\ref{maindef} does not exceed: 
\begin{align}\label{epsdel}
\varepsilon(\delta_1, \delta_2) := \min \left\{ \frac{\delta_1}{ C_1 }, \frac{ \delta_2 L_{2k}(\mathbf{A})}{ 1 + C_1 \left( \max_{i \in [N]} \|\mathbf{x}_i\|_1  + \delta_2 \right) } \right\}.
\end{align}

@@

As already mentioned, it is easy to construct a regular hypergraph $E \subseteq {[m] \choose k}$ with the SIP having cardinality $|E| = m$, implying the lower bound for sample size $N$ from the introduction. In many cases, however, the SIP can be achieved with far fewer supports; for example, when $k = \sqrt{m}$ one can take $E$ to be the $2k$ rows and columns formed by arranging $[m]$ in a square grid.

@@

\begin{corollary}\label{SLCopt}
If all of the assumptions of Thm.\ref{DeterministicUniquenessTheorem} hold with more than $(k-1)\left[ {\bar m \choose k} + |E|k{\bar m \choose k-1}\right]$ vectors $\mathbf{x}_i$ supported on each $S \in E$, then all solutions to Prob.~\ref{OptimizationProblem} necessarily satisfy the implications \eqref{Cstable} and \eqref{b-PDa} of Thm.~\ref{DeterministicUniquenessTheorem}.
\end{corollary}
